% ==============================================================
% UNIVERSITY OF BUCKINGHAM PhD THESIS
% Non-Science | Harvard | 12pt TNR | 1.5 spacing
% Compile: xelatex → biber → xelatex → xelatex
% ==============================================================

\documentclass[12pt,oneside]{report}
\usepackage[a4paper, margin=20mm, top=25mm, bottom=25mm]{geometry}

% Typography
\usepackage{setspace}
\usepackage{fontspec}
\setmainfont{Times New Roman}
\onehalfspacing
\setlength{\parindent}{1.27cm}

% Packages
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{booktabs}
\usepackage[nottoc,numbib]{tocbibind}
\usepackage{csquotes}
\usepackage{appendix}

% Harvard Bibliography
\usepackage[style=authoryear, backend=biber, sorting=nyt]{biblatex}
\addbibresource{refs.bib}

% Page headers
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\slshape\leftmark}
\fancyhead[LO,RE]{\slshape\rightmark}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% Chapter style
\titleformat{\chapter}[display]
  {\normalfont\Large\bfseries\centering}{\chaptertitlename\ \thechapter}{1em}{}
\titlespacing*{\chapter}{0pt}{1.5em}{1em}

\begin{document}

% ========== TITLE PAGE ==========
\begin{titlepage}
\centering
\vspace*{2cm}
{\LARGE \textbf{Corporate Governance and Default Risk in Nigeria’s Deposit Money Banks}\par}
\vspace{1cm}
{\Large A Study of Post-Consolidation Reforms and Regulatory Clarity (2009–2025)\par}
\vspace{2cm}
{\Large Azii Isaiah Jimoh\par}
\vspace{1cm}
{\Large A thesis submitted for the degree of Doctor of Philosophy\par}
\vspace{1cm}
{\Large The University of Buckingham\par}
\vspace{1cm}
{\Large October 2025\par}
\end{titlepage}

% ========== FRONT MATTER (Roman) ==========
\pagenumbering{roman}

\begin{abstract}
\addcontentsline{toc}{chapter}{Abstract}
This thesis investigates corporate governance mechanisms in Nigeria’s deposit money banks and their impact on default risk (measured via non-performing loan ratios) from 2009 to 2024. Using panel fixed-effects models, mediation analysis, and difference-in-differences, it tests four research questions: the influence of governance indices, audit independence mediation via financial reporting quality, disclosure activation post-2014 reforms, and regulatory clarity’s causal effect on NPLs (2020–2025). Findings inform unified, enforceable governance frameworks aligned with Basel III and OECD standards.
\end{abstract}

\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
I express profound gratitude to my supervisor, [Name], for invaluable guidance. I thank the University of Buckingham, Central Bank of Nigeria, and participating bank executives. This work is dedicated to my family.

\tableofcontents

\chapter*{Declaration of Originality and Prior Publication}
\addcontentsline{toc}{chapter}{Declaration of Originality and Prior Publication}
\label{decl-page}

I hereby declare that this thesis is my own original work and has not been submitted elsewhere for a degree. All sources are acknowledged. No part has been previously published except as noted below:

\begin{itemize}
  \item Jimoh, A. I. (2024). ``Audit Independence and NPLs: A Mediation Study.'' \textit{Nigerian Journal of Banking}, under review.
\end{itemize}

% ========== MAIN TEXT (Arabic) ==========
\cleardoublepage
\pagenumbering{arabic}

\input{chapters/chapter01}

% \input{chapters/chapter02} % Add later

% ========== APPENDICES ==========
\begin{appendices}
\chapter{Survey Questionnaire}
Full 7-point Likert scale instrument used in 2024 data collection.
\end{appendices}

% ========== BIBLIOGRAPHY ==========
\cleardoublepage
\printbibliography[title=Bibliography]

\end{document}
